Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-09 text/ms.tex
Copy editing ms

By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process errors have heavy tails and are thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b). For example, at $\nu = 3$, the probability of drawing a value more than five standard deviations below the mean is $0.02$, whereas the probability of drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$). As the value of $\nu$ approaches infinity, the distribution approaches the normal distribution (Fig.~\ref{fig:didactic}a, b).

@@

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}a). Longer time-series length was the strongest covariate of observing black-swan dynamics. For example, we were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a population with 60 time steps compared to one with 30 time steps. However, the absolute change in probability with increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

We examined all time series with published explanations of why the black-swan events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events (Table~\ref{tab:causes-supp}). For example, a shortage of nest sites for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}. The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black-swan event for a water vole (\textit{Arvicola terrestris}) population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in the United Kingdom \citep{stafford1971}. Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}. \textit{(TODO: substitute a parasite example.)}

@@

We found strong evidence for black swans (heavy-tailed process noise) in 3--5\% of ecological time series. Black swans are usually downward events and are detected more frequently in longer time series and in populations with a smaller magnitude of process error. However, black swans were not associated with density dependence, population growth rate, or lifespan. In verified cases, black swans were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures. Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black-swan population dynamics and if anything will underestimate the probability of black swans. The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.

Our results clarify previous related analyses. Previous work with an older version of the GPDD assessed the distribution of abundance time series but focused on identifying if the log-normal distribution was the most frequently parsimonious model \citep{halley2002}. For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series. However, black-swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes. Therefore, we would not expect to observe black swans in a large proportion of populations. By probabilistically quantifying process noise tails on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans. In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance. However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could falsely generate heavy tails. This rebuttal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with the same mean and inverse-gamma-distributed variances,][]{gelman2014}.

We might expect to observe black-swan dynamics in ecological time series because of either unmodelled intrinsic properties of populations or extrinsic forces acting on populations. Since a t-distribution can be formed by a mixture of normal distributions \citep{gelman2014}, we could observe heavy tails if we miss some underlying mixture of intrinsic processes \citep{allen2001}. That process might be an aggregation of populations across space, or population diversity, or some intrinsic change in population variability through time. Extrinsic forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}. These forces could be extreme themselves. For example, extreme climate, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans. Alternatively, the interaction of multiple ``normal'' extrinsic forces could give rise to black-swan ecological dynamics. This could occur if the interaction is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we expect that we would find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans likely differ (e.g.~severe winters in Table~\ref{tab:sparks}). In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. Indeed, three of the populations our method initially identified as heavy tailed turned out to be data-entry errors (Supporting Information). A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions. Ecological resource management can draw from other disciplines that focus on heavy tails. For example, earthquake preparedness and response is focussed on black-swan events. Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans (e.g.~REFERENCE?). A similar focus might benefit resource management once our ability to predict the spatial probability and covariates of ecological black swans improves. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For example, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}. Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when our models break down, and adjust future management policy \citep{doak2008, pine-iii2009, lindenmayer2010}.

Our results suggest important future issues to address about ecological black swans. Given that black swans do occur, can we forecast the probability of black swans in space and time? Furthermore, what management policies allow us to detect them quickly after they happen? Can we isolate the components of ecological dynamics that experience black swans by moving from phenomenological models such as the Gompertz to mechanistic models, that take into account, for example, recruitment? We expect that greater insight into the mechanisms and covariates of ecological black swans may be best obtained through specific geographic and taxonomic subsets of data where longer time series with low levels of observation error are available \citep[e.g.][]{segura2013}.

>>>>

By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b). For example, at $\nu = 3$, the probability of drawing a value more than five standard deviations below the mean is $0.02$, whereas the probability of drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$). As the value of $\nu$ approaches infinity, the distribution approaches the normal distribution (Fig.~\ref{fig:didactic}a, b).

@@

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise ($\sigma$) but not clearly related to population growth rate ($\lambda$), density dependence ($b$), or maximum lifespan (Figs~\ref{fig:correlates}, \ref{fig:posteriors}a). Longer time-series length was the strongest covariate of observing black-swan dynamics. For instance, we were about \pIncHeavyNThirtyNSixty~times as likely to observe heavy tails in a population with 60 time steps compared to one with 30 time steps. However, the absolute change in probability with increased time series length was small ($\pHeavyNSixty$ vs.\ $\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

We examined all time series with published explanations of why the black-swan events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The majority (\percBSDown \%) of documented events were downward black swans and involved a series of unfortunate events (Table~\ref{tab:causes-supp}). For example, a shortage of nest sites for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}. The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black-swan event for a water vole (\textit{Arvicola terrestris}) population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in the United Kingdom \citep{stafford1971}. Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}. \textit{(TODO: substitute a parasite example.)}

@@

We found strong evidence for black swans (heavy-tailed process noise) in 3--5\% of ecological time series. Black swans were usually downward events and were detected more frequently in longer time series and in populations with a smaller magnitude of process noise. Black swans were not associated with density dependence, population growth rate, or lifespan. In verified cases, black-swan events were often a result of the interaction between elements of extreme climate, predators and parasite cycles, and strong changes in human pressures. Our empirical results, sensitivity analyses, and simulation tests suggest that estimating the tail shape of process noise is a viable method of detecting black-swan population dynamics and if anything will underestimate the probability of black-swan events. The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century \citep{easterling2000,ipcc2012}.

Our results clarify previous related analyses. An analysis with an older version of the GPDD assessed the distribution of abundance time series but focused on identifying if the log-normal distribution was the most frequently parsimonious model \citep{halley2002}. For heavy-tailed distributions, \citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the four-parameter Levy stable distribution and found little information criteria support for these distributions in longer time series. However, black-swan events are by definition rare, and the majority of time series in the GPDD are short for these purposes. Therefore, we would not expect to observe black swans in a large proportion of populations. By probabilistically quantifying process noise tails on a continuous spectrum, and allowing for non-stationary time series and density dependence, our analysis allows for a more nuanced description of the evidence of ecological black swans. In an earlier study, \citet{keitt1998} described heavy (power law) tails in breeding bird population abundance. However, this finding was challenged by \citet{allen2001}, who showed that mixing data across species could falsely generate heavy tails. This rebuttal supports our use of the t distribution to represent heavy tails, since a t distribution can be represented by a mixture of normal distributions \citep[with the same mean and inverse-gamma-distributed variances,][]{gelman2014}.

We might expect to observe black-swan dynamics in ecological time series because of unmodelled intrinsic properties of populations or extrinsic forces acting on populations. Since a t-distribution can be formed by a mixture of normal distributions \citep{gelman2014}, we could observe heavy tails if we miss some underlying mixture of intrinsic processes \citep{allen2001}. That process might be an aggregation of populations across space, or population diversity, or an intrinsic change in population variability through time. Extrinsic forces could also cause black-swan dynamics \citep[e.g.][]{nunez2012}. These forces could be extreme themselves. For example, extreme climate, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressure such hunting, fishing, or habitat destruction might cause black swans. Alternatively, the interaction of multiple ``normal'' extrinsic forces could give rise to black-swan ecological dynamics. This could occur if the interaction is synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions experience a rare alignment \citep{denny2009}.

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we would expect to find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ (e.g.~severe winters in Table~\ref{tab:sparks}). In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. Indeed, three of the populations our method initially identified as heavy tailed turned out to be data-entry errors and were discarded (Supporting Information). A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. As an example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions. Ecological resource management can draw from other disciplines that focus on heavy tails. For example, earthquake preparedness and response is focussed on black-swan events. Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans. A similar focus might benefit resource management once our ability to predict the spatial probability and covariates of ecological black swans improves. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For instance, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events \citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations persist when others are affected by disease or extreme environmental forces \citep[e.g.][]{hilborn2003, schindler2010, anderson2014}. Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when our models break down, and adjust future management policy \citep{doak2008, pine-iii2009, lindenmayer2010}.

Our results suggest a number of research questions related to ecological black swans. Given that black swans do occur, can we forecast the probability of black swans in space and time? Furthermore, what management policies allow us to detect them quickly after they happen? Can we isolate the components of ecological dynamics that experience black swans by moving from phenomenological models such as the Gompertz to mechanistic models, that take into account, for example, recruitment? We expect that greater insight into the mechanisms and covariates of ecological black swans may be best obtained through specific geographic and taxonomic subsets of data where longer time series with low levels of observation error are available \citep[e.g.][]{segura2013}.
